---
title: "Phospho (STY) sites analysis"
author: "Stephen Shoebridge, stephen.shoebridge@univie.ac.at"
date created: "16/12/2021"
updated date: "18/01/2023"
output: 
  html_document:
    self-contained: TRUE
---

```{r setup, include=FALSE, echo=FALSE}
library(tidyverse)
library(plotly)
library(limma)
library(readxl)
library(data.table)
library(ggbreak)
library(ggpubr)
```

# Introduction

In this analysis we analyse the mass spectrometer data provided. We read in a trimmed version of the data and establish the following cut-offs

```{r read files, include=T}
phospho_data <- read_xlsx("/Users/stephen/Library/CloudStorage/OneDrive-Personal/Martens Lab Onedrive/Bioinformatics/PI3K Phosphorylation/LCMS_report_trimmed.xlsx", sheet = 1)#Take peptide_intensity of Phospho sites
colnames(phospho_data)#Names of each column
metadata <- column_to_rownames(read_xlsx("/Users/stephen/Library/CloudStorage/OneDrive-Personal/Martens Lab Onedrive/Bioinformatics/PI3K Phosphorylation/Phosphosite_metadata.xlsx"), "Label")
head(metadata)#Labels broken down into data fields
```

The data frame contains the peptide intensity for the TBK1 and ULK1 samples. The metadata contains a breakdown of the sample labels.

# Cutoffs

In the phospho site data, the following cut-offs will be applied

-   If the sum of the row of the peptide is equal to 0, it will be excluded

-   If the peptides score is less than 90 it is excluded

-   All samples with a score between 90-100 are labelled with a low-score warning in the metadata

```{r Cutoffs, echo=F}
phospho_sel <- subset(phospho_data, rowSums(phospho_data[, 19:46])>0)
phospho_sel <- subset(phospho_sel, phospho_sel$Score>90)
paste0("Number of rows before cutoff: ", nrow(phospho_data), "     Number of rows after cutoff: ", nrow(phospho_sel))
phospho_sel$Low_Score_Warning <- phospho_sel$Score<100
table(phospho_sel$Low_Score_Warning); print("7 peptides have a score between 90 and 100")
phospho_sel <- subset(phospho_sel, phospho_sel$Comments=="OK")#3 removed with bad comments
```

## Subtracting the baseline

```{r Subtracting tine point 0 from time point 30}
phospho_sel$pep_label <- 1:nrow(phospho_sel)
phospho_mat <- column_to_rownames(phospho_sel[,c(19:46, (ncol(phospho_sel)))], "pep_label")#Gives the rows an identifier since no column iD seems unique
phospho_sel <- column_to_rownames(phospho_sel, "pep_label")#Same for phospho_sel
metadata$label <- paste0(metadata$User, "_",  metadata$Target, "_",  metadata$Band)
metadata <- metadata[order(metadata$label, metadata$Time),]
metadata$sample_name <- rownames(metadata)
timepoint_0 <- subset(metadata, metadata$Time==0)$sample_name
timepoint_30 <- subset(metadata, metadata$Time==30)$sample_name
timepoint_0_mat <- phospho_mat[, timepoint_0]
timepoint_30_mat <- phospho_mat[, timepoint_30]
setnames(timepoint_0_mat, old = subset(metadata, metadata$Time=="0")$sample_name, new = subset(metadata, metadata$Time=="0")$label)
setnames(timepoint_30_mat, old = subset(metadata, metadata$Time=="30")$sample_name, new = subset(metadata, metadata$Time=="30")$label)
cols <- sort(intersect(names(timepoint_0_mat), names(timepoint_30_mat)))
setdiff(names(timepoint_0_mat), names(timepoint_30_mat))#Check that no columns have non matching titles
substr_mat <- timepoint_30_mat[cols]-timepoint_0_mat[cols]
substr_mat[substr_mat<0] <- NA # Ignores negative values
phospho_subtr <- cbind(phospho_sel[, c(1:18, 47)], substr_mat)
```

## Write the cleaned file to csv for later reference

```{r}

#write.csv(phospho_subtr, "~/OneDrive/Martens Lab/R Work/phospho_sites_30sub0.csv")
colnames(phospho_subtr)
phospho_subtr_sel <- subset(phospho_subtr, rowSums(phospho_subtr[, 21:33], na.rm = T)>0)#ignores NA and removes rows that have a sum of 0
nrow(phospho_subtr); nrow(phospho_subtr_sel)
#phospho_subtr <- read.csv("~/OneDrive/Martens Lab/R Work/phospho_sites_30sub0.csv")
#write.csv(phospho_subtr_sel, "~/OneDrive/Martens Lab/R Work/phospho_sites_30sub0.csv")
```

# Plots

## Interactive graphs of the dataset

```{r}
melt_phospho <- melt(phospho_subtr[,c(1,3,6,7,20:33)], id.vars=c("Proteins", "Positions within proteins", "Protein names", "Gene names"), na.rm = T)
melt_phospho <- melt_phospho[order(melt_phospho$`Positions within proteins`),]
melt_phospho$`Positions within proteins` <- gsub(".*;", "", melt_phospho$`Positions within proteins`)#removes N terminal before semi colon (position)
melt_phospho$`Positions within proteins` <- as.numeric(melt_phospho$`Positions within proteins`)#turns the numbers to numeric
melt_phospho <- melt_phospho %>% dplyr::arrange(`Positions within proteins`)
melt_phospho <- melt_phospho[order(as.numeric(melt_phospho$`Positions within proteins`)),]
melt_phospho$`Gene names` <- ifelse(is.na(melt_phospho$`Gene names`), melt_phospho$Proteins, melt_phospho$`Gene names`)
melt_phospho$log10PI<- log10(melt_phospho$value+1)
p <- ggplot(melt_phospho, aes(x=`Positions within proteins`, y=log10PI, color=variable))+geom_point()+facet_wrap(~`Gene names`, scales = "free_x")+theme(axis.text.x = element_text(angle = 90))+
  labs(title = "Plot of position of peptide on protein and peptide intensity", x="Position on Protein", y="Log10(Difference in Peptide Intensity, (t30-t0))")+scale_x_continuous()+scale_y_continuous(limits = c(0,10))
pg <- ggplotly(p)
pg
p
```

## Publication plots

For the publication several plots are made for the figure. For this the mCherry_Q6ZNE5 can be removed. In addition plots should take into account whether the protein is from the correct band or is just co-selected in the same band. Positions within the protein are labelled in each plot. TBK1 and ULK1 are separated by colour.

### Phosphorylation site regardless of bands

```{r}
melt_phospho[c('User', 'Kinase', "Band")] <- str_split_fixed(melt_phospho$variable, '_', 3)
phospho_sel <- subset(melt_phospho, `Gene names`!="mCherry_Q6ZNE5")#Remove mCherry identification
#Changing alias names
phospho_sel$target <- ""
phospho_sel[phospho_sel$`Gene names`=="PIK3R4", "target"] <- "VPS15"
phospho_sel[phospho_sel$`Gene names`=="PIK3C3", "target"] <- "VPS34"
phospho_sel[phospho_sel$`Gene names`=="BECN1", "target"] <- "Beclin1"
phospho_sel[phospho_sel$`Gene names`=="ATG14", "target"] <- "ATG14"

phospho_sites <- phospho_sel %>% group_by(`Positions within proteins`, target) %>%
  summarise(max = max(log10PI, na.rm=TRUE))


p1 <- ggplot(phospho_sel, aes(x=`Positions within proteins`, y=log10PI))+
  geom_col(width=0.1, color="black", position = position_identity())+
  geom_point(shape=21, aes(fill=Kinase), size=4)+
  geom_text(data=phospho_sites, aes(x=`Positions within proteins`, y=max+0.7, label=`Positions within proteins`), angle=90)+
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA))+
  scale_x_continuous(breaks = scales::pretty_breaks())+
  theme_bw()+
  labs(y="Log10 Adjusted Peptide Intensity (30min-0min)", title = "Band≠Target")+
  facet_wrap(~target, ncol = 2, scales = "free_x")
```

## Removing proteins that are not found in their appropriate band

```{r}
melt_phospho[c('User', 'Kinase', "Band")] <- str_split_fixed(melt_phospho$variable, '_', 3)
phospho_sel <- subset(melt_phospho, `Gene names`!="mCherry_Q6ZNE5")#Remove mCherry identification
#Changing alias names
phospho_sel$target <- ""
phospho_sel[phospho_sel$`Gene names`=="PIK3R4", "target"] <- "VPS15"
phospho_sel[phospho_sel$`Gene names`=="PIK3C3", "target"] <- "VPS34"
phospho_sel[phospho_sel$`Gene names`=="BECN1", "target"] <- "Beclin1"
phospho_sel[phospho_sel$`Gene names`=="ATG14", "target"] <- "ATG14"

#Band=target
phospho_sel <- phospho_sel[phospho_sel$target==phospho_sel$Band,]

phospho_sites <- phospho_sel %>% group_by(`Positions within proteins`, target) %>%
  summarise(max = max(log10PI, na.rm=TRUE))


p2 <- ggplot(phospho_sel, aes(x=`Positions within proteins`, y=log10PI))+
  geom_col(width=0.1, color="black", position = position_identity())+
  geom_point(shape=21, aes(fill=Kinase), size=4)+
  geom_text(data=phospho_sites, aes(x=`Positions within proteins`, y=max+0.7, label=`Positions within proteins`), angle=90)+
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA))+
  scale_x_continuous(labels = scales::label_number(accuracy = 1), breaks = scales::breaks_pretty())+
  theme_bw()+
  labs(y="Log10 Adjusted Peptide Intensity (30min-0min)", title = "Band=Target")+
  facet_wrap(~target, ncol = 2, scales = "free_x")
p2
```

```{r}
ggarrange(p1, p2, ncol=1)
```
